Reinforcement Learning Produces Dominant Strategies for the Iterated Prisoner's Dilemma

### 2017-07-04 main.tex
Update section on Noisy tournaments.

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./plots/EvolvedLookerUp2_2_2.png}
        \caption{Fingerprint for EvolvedLookerUp\_2\_2\_2}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./plots/Evolved_HMM_5.png}
        \caption{Fingerprint for Evolved\_HMM\_5}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./plots/Evolved_FSM_16.png}
        \caption{Fingerprint for Evolved\_FSM\_16}
    \end{subfigure}%

@@

We also ran noisy tournaments in which there is a 10\% chance that an action
is flipped. The best performing strategies in mean payoff are DBS, designed
to correct for noise, followed by two strategies trained
in the presence of noise, Spiteful TFT (TFT but defects indefinitely if the
opponent defects twice consecutively), two strategies trained without noise, then
OmegaTFT (also designed to handle noise).

The strategies trained in the presence of noise are also among the best performers
in the absence of noise. The cluster of mutually cooperative strategies is
broken by the noise at 5\%. A similar collection of players excels at winning
matches but again they have a poor total payoff.

The strategies tallying the most wins are somewhat similar, with Defector, the
handshaking CollectiveStrategy, and Aggravate appearing as the top three again,
followed by Spiteful TFT and strategies designed to be retaliatory. Players ranked
12, 13, and 15 (TF3, Fool Me Once, and Evolved ANN 5) all appear in the top 15
strategies by score, indicating that noise breaks the asymmetry between wins
and scores to some extent.

% Table of best strategies

\begin{table}[!hbtp]
    \centering
        \input{./tex/noisy_scores.tex}
        \caption{Noisy Tournament: Top Ranking Strategies by Median Score in a
        15000 Tournaments with 10\% noise}
        \label{tbl:noisy_score}
\end{table}

\begin{landscape}
    \begin{figure}[!hbtp]
        \includegraphics[width=\paperwidth]{plots/Noisy_boxplot.pdf}
        \caption{Ranking by Median Payoff for the Noisy Tournament}
    \end{figure}
\end{landscape}

\begin{landscape}
    \begin{figure}[!hbtp]
        \includegraphics[width=\paperwidth]{plots/Noisy_winplot.pdf}
        \caption{Ranking by Total wins for the Noisy Tournament}
    \end{figure}
\end{landscape}

\begin{figure}[!hbtp]
    \includegraphics[width=\paperwidth]{plots/Noisy_payoff.pdf}
    \caption{Pairwise Match results for the Noisy Tournament}
\end{figure}

@@

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./plots/DBS.png}
        \caption{Fingerprint for DBS}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./plots/Evolved_ANN_5_Noise_05.png}
        \caption{Fingerprint for Evolved\_ANN\_5\_Noise\_05}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./plots/Evolved_FSM_16_Noise_05.png}
        \caption{Fingerprint for Evolved\_FSM\_16\_Noise\_05}
    \end{subfigure}%

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. A simple evolutionary algorithm produces strategies based on multiple
standard machine learning techniques that are able to achieve a higher mean
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies (though the trained memory one strategy performs
well). The generic structure of the trained strategy did not appear to be
critical -- strategies based on lookup tables, finite state machines, and stochastic
variants all performed well. Single layer neural networks also performed well
though these had some aspect of human involvement in the selection of features.
The success of the other strategy types suggests that a deeper network that
incorporates feature engineering would likely also perform well.

@@

Of the strategies trained to maximize their mean score all are generally
cooperative, not defecting until the opponent defects. Maximizing for individual
performance across a collection of opponents leads to mutual cooperation despite
the fact that mutual cooperation is an unstable equilibrium for the prisoner's
dilemma. Specifically we note that the reinforcement learning process for maximizing
payout does not
lead to exploitative zero determinant strategies, which may be a result of the
collection of training strategies, many of which retaliate harshly.

>>>>

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./assets/EvolvedLookerUp2_2_2.png}
        \caption{Fingerprint for EvolvedLookerUp\_2\_2\_2}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./assets/Evolved_HMM_5.png}
        \caption{Fingerprint for Evolved\_HMM\_5}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./assets/Evolved_FSM_16.png}
        \caption{Fingerprint for Evolved\_FSM\_16}
    \end{subfigure}%

@@

We also ran noisy tournaments in which there is a 5\% chance that an action is
flipped. As shown in Table~\ref{tbl:noisy_score} and
Figure~\ref{fig:noisy_score}, the best performing strategies in median payoff
are DBS, designed to correct for noise, followed by two strategies trained in
the presence of noise and three trained strategies trained without noise. One of
the strategies trained with noise (PSO Gambler) actually performs less well than
some of the other high ranking strategies including
Spiteful TFT (TFT but defects indefinitely if the opponent defects twice
consecutively) and OmegaTFT (also
designed to handle noise).

% Table of best strategies
\begin{table}[!hbtp]
    \centering
        \input{./assets/noisy_top_15_scores_summary.tex}
        \caption{Noisy Tournament: Top Ranking Strategies by Median Score in a
        15000 Tournaments with 5\% noise}
        \label{tbl:noisy_score}
\end{table}

\begin{landscape}
    \begin{figure}[!hbtp]
        \includegraphics[width=\paperwidth]{./assets/noisy_scores_boxplots.pdf}
        \caption{Ranking by Median Payoff for the Noisy Tournament}
        \label{fig:noisy_score}
    \end{figure}
\end{landscape}

The strategies trained in the presence of noise are also among the best
performers in the absence of noise. As shown in Figure~\ref{fig:noisy_heatmap}
the cluster of mutually cooperative strategies is broken by the noise at 5\%. A
similar collection of players excels at winning matches but again they have a
poor total payoff.

\begin{figure}[!hbtp]
    \includegraphics[width=\textwidth]{./assets/noisy_scores_heatmap.pdf}
    \caption{Pairwise Match results for the Noisy Tournament}
    \label{fig:noisy_heatmap}
\end{figure}

As shown in Figure~\ref{fig:noisy_winplot} the strategies tallying the most wins
are somewhat similar, with Defector, the handshaking CollectiveStrategy, and
Aggravate appearing as the top three again.

\begin{landscape}
    \begin{figure}[!hbtp]
        \centering
        \includegraphics[width=\textwidth]{./assets/noisy_wins_boxplots.pdf}
        \caption{Ranking by Total wins for the Noisy Tournament}
        \label{fig:noisy_winplot}
    \end{figure}
\end{landscape}

As shown in Table~\ref{tbl:noisy_wins}, the top ranking strategies win a larger
number of matches in the presence of noise. For example Spiteful Tit For Tat in
one tournament won almost all it's matches (167).

\begin{table}[!hbtp]
    \centering
        \input{./assets/noisy_top_15_wins_summary.tex}
        \caption{Noisy Tournament: Number of wins of Top Ranking Strategies
        by median score (50th percentile) in 15,000 Tournaments}
        % TODO Update caption as more repetitions are carried out
        % TODO Potentially remove count from table (including it for now so as
        % to avoid confusion.
        \label{tbl:noisy_wins}
\end{table}

Finally, Table~\ref{tbl:noisy_ranks} and
Figure~\ref{fig:noisy_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. We see, that the stochasticity
of the ranks understandably increases the DBS strategy never ranks lower than
third and wins 75\% of the time. The two strategies trained for noisy
tournaments rank in the top three 95\% of the time.
% TODO Possibly update this with more data.

\begin{table}[!hbtp]
    \centering
        \input{./assets/noisy_top_15_ranks_summary.tex}
        \caption{Noisy Tournament: Ranks of Top Ranking Strategies
        by median score (50th percentile) in 15,000 Tournaments}
        % TODO Update caption as more repetitions are carried out
        % TODO Potentially remove count from table (including it for now so as
        % to avoid confusion.
        \label{tbl:noisy_ranks}
\end{table}

\begin{landscape}
    \begin{figure}[!hbtp]
        \centering
        \includegraphics[width=\textwidth]{./assets/noisy_ranks_boxplots.pdf}
        \caption{Rankings over each repeated tournament}
        \label{fig:noisy_ranks_boxplot}
    \end{figure}
\end{landscape}

@@

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./assets/DBS.png}
        \caption{Fingerprint for DBS}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./assets/Evolved_ANN_5_Noise_05.png}
        \caption{Fingerprint for Evolved\_ANN\_5\_Noise\_05}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{./assets/Evolved_FSM_16_Noise_05.png}
        \caption{Fingerprint for Evolved\_FSM\_16\_Noise\_05}
    \end{subfigure}%

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. A simple evolutionary algorithm produces strategies based on multiple
standard machine learning techniques that are able to achieve a higher median
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies (though the trained memory one strategy performs
well). The generic structure of the trained strategy did not appear to be
critical -- strategies based on lookup tables, finite state machines, and stochastic
variants all performed well. Single layer neural networks also performed well
though these had some aspect of human involvement in the selection of features.
The success of the other strategy types suggests that a deeper network that
incorporates feature engineering would likely also perform well.

@@

Of the strategies trained to maximize their median score all are generally
cooperative, not defecting until the opponent defects. Maximizing for individual
performance across a collection of opponents leads to mutual cooperation despite
the fact that mutual cooperation is an unstable equilibrium for the prisoner's
dilemma. Specifically we note that the reinforcement learning process for maximizing
payout does not
lead to exploitative zero determinant strategies, which may be a result of the
collection of training strategies, many of which retaliate harshly.
